\documentclass[10pt, a4paper]{article}

\input{./template/template.tex}

\begin{document}

\section*{Software and Security Engineering Supervision 3}

There is a list of past exam questions and relevant courses on the 2021 course pages.

\section{Leaks}

\begin{itemize}

\item The question would want you to state that the requirements were convoluted.

\item Accidents are the most likely way of leaking information.

\item You should have gatekeepers (tests) which it is not possible to bypass.

You would use proxy tools to gatekeep. Have multiple people set these.
For example setting a time of release.

\item Making people work on a specific machine in an inaccessible room would
remove the chance of accidental error.

\item Consider insider threats.

\end{itemize}

\section{Deterrent}

\begin{itemize}

\item Make threats tamper evident. If someone steals an exam question and you are
guaranteed that you can change it then you know they won't bother.

\item ``Blame and train'' is completely the wrong approach -- if you
set the rules to be stupid then people will subvert them and the only
solution is to make a policy which is easier to use.

\item Certain corporates will deliberately phish all of your staff to try and find out
who will get phished. This is very bad and you end up blame and train and you end
up getting noncompliance elsewhere.

\end{itemize}

\section{Agile Development}

\begin{itemize}

\item If you want to work out which strategy is best then you need to figure out which
strategy is best. This is usually having short deadlines. However in some cases this
comes across as coercive. Good project management is about figuring out which is
best for different people and managing them like that.

\item Agile suggests 2 week deadlines because most people will only work intensively in the
last week. You want to balance the stress of short deadlines with getting things
done.

If you can't split a task up then you will ``be screwed'' -- you have to split tasks
up as much as you can -- it makes things more achievable.

Work on your worst problem first. If you can't solve the worst problem then
you should scrap the project.

Even if you're not being told to solve your worst problem then you should still
do it.

\item You should do PERT charts to find out the critical paths and then
make GANTT charts of the critical paths. GANTT are for managing people and for
management.

\item If anything on the critical path gets delayed then it will delay the whole
project. You use the slack time on certain paths to decide which people
to use and when.

\item Automated regression testing checks you don't reintroduce bugs. People are
likely to make the same bugs because they're intrinsically complex and fixes
are often brittle.

\item Automatedd Regression Testing is a \textit{part} of Continuous Integration!
You will run automated tests (automated regression tests, unit tests and system tests).

\item In nightlies (nightly builds), you run a ton of tests and build. If it fails
some tests then you roll back to the last day it worked. If it passes then you release
a nightly build.

\item What people say in scums are usually diffs based on the previous day.
What they say is very similar to the days before. The message that comes out
each day is nonsense. It's motivational to know that you've got to tell
people what you've been doing.

Managers aren't allowed to talk in scrums. Systems such as agile allow you to
subvert managers -- it's not a great way to run a company if you've got good managers.
It forces managers to listen.

If you have a problem and say it in a scrum then you may find other people who
have found the same problem and can help you.

\item Git is a usability disaster although version control is necessary.

\end{itemize}

\section{Protocols}

\begin{itemize}

\item Timestamps don't need to be physical timestamps -- they can be logical timestamps
(which are just counters).

\item Consider man-in-the-middle attacks and consider brute force attacks!
Messages which are signed need to have the sender/receivers ID. You can usually
interpret timestamps to be nonces.

\end{itemize}

\end{document}
