\documentclass[10pt, a4paper]{article}

\input{./template/template.tex}

\begin{document}

\section*{MLRWD Session notes}

You want to find the \textbf{features} which makes the thing the most accurate.

\[
\hat{c} = \text{argmax}_{c \in C} \log P(c) + \sum_i \log P(f_i|c)
\]
Choose the most appropriate $f_i$! This may either be individual objects or the
presence or absence of a type.

Often considering a feature to be the presence or absence of a type.

Mention testing things empirically.

Consider the presence or absence of types of trees. It's often more appropriate to measure
absence when there is a small number of types of which most are almost always present.

\section{Cross-Fold Validation}

Cross-fold-validation is a way of enhancing the training data you do have
by ``re-using'' it.

There are two types of cross-fold validation: stratified and sampled.

Stratified is when you make sure the proportion of each class in each fold is approximately right.

We should always use stratified cross validation. Random cross-fold validation ends up
being pretty poor.

\section{Testing}

The testing process involves evaluation! Make sure to mention it in questions discussing
testing.

\section{Read the question!}

2018P3Q7e:

Build a lexicon classifier and use a combination of this classifier and the
na\"{\i}ve bayes classifier to predict the actual result.

Use semi-supervised learning with the sentiment classifier to label data for the na\"{\i}ve bayes
classifier.

Smoothing parameter. If I don't see a tree species with a class in my data and I also don't see it in the dictionary
then I assume it will never appear (this stops shifting probability mass and gives better parameters).

Something to do with ``tuning some parameters'' -- doing something in terms of development.

The only parameter to smooth is the smoothing parameter.

Significance tests! Sign test! Relearn the formula!

Justify which you should use -- one tailed or two tailed!

\section{Parameters}

Parameter estimation is \textbf{going to be simple} -- absolutely do not overcomplicate things!

\end{document}
