\input{./infofile.tex}

\documentclass[10pt,\jkfside,a4paper]{article}

\input{../../template/template.tex}

\begin{document}

\begin{examquestion}{2007}{4}{1}

\begin{enumerate}[label=(\alph*)]

\item What is meant by a \textit{serialisable} order for two or more
transactions?

% TODO

\item Explain how \texit{timestamp ordering} enforces isolation.

% TODO

\item Draw and explain a history graph for two transactions whose invocations
of a set of conflicting operations are serialisable but who are rejected by
TSO .

% TODO

\item Considering Optimistic Concurrency Control (OCC):

\begin{enumerate}[label=(\roman*)]

\item State the properties of a transaction's set of shadow copies that must
be verified at commit time.

% TODO

\item Carefully explain the algorithm used by a single-threaded commit-time
validator.

% TODO

\end{enumerate}

\item Consider a system in which the transactions cause updates to objects
which are not all located on a single server but which are distributed
widely around the Internet. What factors would influence your choice of
using TSO or OCC to enforce isolation?

% TODO

\end{enumerate}

\end{examquestion}

\begin{examquestion}{2011}{5}{7}

Database concurrency control enables multiple transactions to proceed
simultaneously against a shared database.

\begin{enumerate}[label=(\alph*)]

\item Give an example of \textit{conflicting} and \textit{non-conflicting}
operations on objects in a database.

% TODO

\item Explain how timestamp ordering (TSO) works, and how it responds to
conflicts between transactions.

% TODO

\item Explain cascading aborts, and use an example to show how they can
occur with TSO .

% TODO

\item Why is TSO not subject to deadlock?

% TODO

\item Explain why TSO distributes well, compared with two-phase locking.

% TODO

\item Construct a workload that performs badly under TSO, and explain why it
does so. What could you do to improve its performance.

% TODO

\end{enumerate}

\end{examquestion}

\end{document}
