\input{./template/prefs/jkf21.tex}
\input{./template/prefs/me.tex}

\documentclass[10pt,\jkfside,a4paper]{article}

\input{./template/includes.tex}
\input{./template/template.tex}

\usepackage{physics}

\begin{document}

Taking the \textbf{differential} (not the derivative) gives:
\begin{equation}
\dd{\ln y} = \frac{\dd{y}}{y} \\
\end{equation}
This is the differential of $\ln y$ -- not with respect to anything but just taking the differential.

To do S5 we will take differentials of this.
\begin{equation}
\begin{split}
f(x, y, z) &= xyz + x^3 + y^4 + z^5 \\
\dd{f} &= \dd{xyz} + \dd{x^3} + \dd{y^4} + \dd{z^5} \\
\dd{f} &= yz\dd{x} + xz\dd{y} + xy\dd{z} + 3x^2\dd{x} + 4y^3\dd{y} + 5z^4\dd{z} \\
\dd{f} &= (yz + 3x^2)\dd{x} + (xz + 4y^3)\dd{y} + (xy + 5z^4)\dd{z} \\
\end{split}
\end{equation}

The partial derivative with some constant kept constant is equal to the normal derivative with the 
other variable kept constant. It is the exact same. Dividing the differential can give easier methods 
for finding this:

\begin{equation}
\begin{split}
\left(\pdv{x}{y}\right)_y &\equiv \left(\dv{x}{y}\right)_z \\
0 &= (yz + 3x^2)\dv{x}{y} + (xz + 4y^3) + (xy + 5z^4)\left(\dv{z}{y}\right)_z \\
0 &= (yz + 3x^2)\dv{x}{y} + (xz + 4y^3) \\
\left(\dv{x}{y}\right)_z &= -\frac{xz + 4y^3}{yz + 3x^2} \\
\left(\pdv{x}{y}\right)_z &= -\frac{xz + 4y^3}{yz + 3x^2} \\
\end{split}
\end{equation}
Note that $\dd{z}{y}$ with $z$ held constant is 0.

Consider a function $f = f(x, y, z) = 0$.

\begin{equation}
\begin{split}
\dd{f} &= \left(\pdv{f}{x}\right)_{y,z}\dd{x} + \left(\pdv{f}{y}\right)_{x,z}\dd{y} + \left(\pdv{f}{z}\right)_{x,y}\dd{z} \\
0 &= \left(\pdv{f}{x}\right)_{y,z}\dd{x} + \left(\pdv{f}{y}\right)_{x,z}\dd{y} + \left(\pdv{f}{z}\right)_{x,y}\dd{z} \\
0 &= \left(\pdv{f}{x}\right)_{y,z}\left(\pdv{x}{y}\right)_z + \left(\pdv{f}{y}\right)_{x,z}\left(\dd{y}{y}\right)_{z} + \left(\pdv{f}{z}\right)_{x,y}\left(\pdv{z}{y}\right)_z \\
0 &= \left(\pdv{f}{x}\right)_{y,z}\left(\pdv{x}{y}\right)_z + \left(\pdv{f}{y}\right)_{x,z} \times 1 + \left(\pdv{f}{z}\right)_{x,y}\times 0 \\
-\left(\pdv{f}{y}\right)_{x,z} &= \left(\pdv{f}{x}\right)_{y,z}\left(\pdv{x}{y}\right)_z \\
\left(\pdv{x}{y}\right)_z &= -\frac{\left(\pdv{f}{y}\right)_{x,z}}{\left(\pdv{f}{x}\right)_{y,z}} \\
\end{split}
\end{equation}
For the other two functions:
\begin{equation}
\begin{split}
\left(\pdv{y}{z}\right)_x &= -\frac{\left(\pdv{f}{z}\right)_{x,y}}{\left(\pdv{f}{y}\right)_{x,z}} \\
\left(\pdv{z}{x}\right)_y &= -\frac{\left(\pdv{f}{x}\right)_{y,z}}{\left(\pdv{f}{z}\right)_{x,y}} \\
\end{split}
\end{equation}

Repeating this for other functions gives similar results. All the denominators and numerators of the functions 
cancel out. This gives:
\begin{equation}
\left(\pdv{x}{y}\right)_x\cdot\left(\pdv{y}{z}\right)_x\cdot\left(\pdv{z}{x}\right)_y = -1 \\
\end{equation}

This generalises to functions with higher numbers of variables: if you multiply the partial derivatives of a 
$n$ variable function together then it is equal to $(-1)^n$.\\
IE for any arbitrary function $f$:
\begin{equation}
f = f(x_1, x_2, x_3, x_4, x_5) \\
\end{equation}
\begin{equation}
\begin{split}
\left(\pdv{x_1}{x_2}\right)_{x_3, x_4, x_5} \times \left(\pdv{x_2}{x_3}\right)_{x_1, x_4, x_5} \times \left(\pdv{x_3}{x_4}\right)_{x_1, x_2, x_5} \times \left(\pdv{x_4}{x_5}\right)_{x_1, x_2, x_3} \times \left(\pdv{x_5}{x_1}\right)_{x_2, x_3, x_4} &= (-1)^5 \\
\left(\pdv{x_1}{x_2}\right)_{x_3, x_4, x_5} \times \left(\pdv{x_2}{x_3}\right)_{x_1, x_4, x_5} \times \left(\pdv{x_3}{x_4}\right)_{x_1, x_2, x_5} \times \left(\pdv{x_4}{x_5}\right)_{x_1, x_2, x_3} \times \left(\pdv{x_5}{x_1}\right)_{x_2, x_3, x_4} &= -1 \\
\end{split}
\end{equation}

Also note that when you partially derive a $n$ variable function then you need to keep $n - 2$ variables constant. 

Ie if $f = f(x_1\dots x_n)$

Then the partial derivative of $x_1$ with respect to $x_2$ is:
\begin{equation}
\left(\pdv{x_1}{x_2}\right)_{x_3\dots x_n}
\end{equation}

Once you have the first derivative of a function then you can use the operator method to find the second 
derivative very quickly as shown below.

\begin{equation}
\begin{split}
x &= u\cos\theta - v\sin\theta \\
y &= u\sin\theta + v\cos\theta \\
\end{split}
\end{equation}
\begin{equation}
\begin{split}
\pdv{f}{u} = \left(\pdv{u}\right) f &= \pdv{f}{x}\pdv{x}{u} + \pdv{f}{y}\pdv{y}{u} \\
\pdv{f}{u} = \left(\pdv{u}\right) f &= \pdv{f}{x}\cos\theta + \pdv{f}{y}\sin\theta \\
\pdv{f}{v} \left(\pdv{v}\right) f &= \pdv{f}{x}\pdv{u}{v} + \pdv{f}{y}\pdv{y}{u} \\
\pdv{f}{v} = \left(\pdv{v}\right) f &= -\pdv{f}{x}\sin\theta + \pdv{f}{y}\cos\theta \\
\end{split}
\end{equation}

So we know what the differential operator with respect to $u$ does to a function. This means 
we can re-apply this to $\pdv{f}{u}$ immediately giving us a function for $\pdv[2]{f}{u}$ 
without further calculation.

\begin{equation}
\begin{split}
\pdv[2]{f}{u} = \left(\pdv{u}\right)\pdv{f}{u} &= \pdv[2]{f}{x}\cos^2\theta + \pdv[2]{f}{y}\sin^2\theta + 2\pdv[2]{f}{x}{y}\sin\theta\cos\theta \\
\pdv[2]{f}{v} = \left(\pdv{v}\right)\pdv{f}{v} &= \pdv[2]{f}{x}\sin^2\theta + \pdv[2]{f}{y}\cos^2\theta - 2\pdv[2]{f}{x}{y}\sin\theta\cos\theta \\
\end{split}
\end{equation}

\begin{equation}
\begin{split}
\pdv[2]{f}{u} + \pdv[2]{f}{v} &= (\sin^2\theta + \cos^2\theta)\pdv[2]{f}{x} + (\sin^2\theta + \cos^2\theta)\pdv[2]{f}{y} \\
\pdv[2]{f}{u} + \pdv[2]{f}{v} &= \pdv[2]{f}{x} + \pdv[2]{f}{y} \\
\end{split}
\end{equation}

\end{document}