\input{./template/prefs/jkf21.tex}
\input{./template/prefs/me.tex}

\documentclass[10pt,\jkfside,a4paper]{article}

\input{./template/includes.tex}
\input{./template/template.tex}

\usepackage{physics}

\begin{document}

Using operators to solve second order ODE's:
consider
\begin{equation}
\dv[2]{y}{x} - 5\dv{y}{x} + 6y = e^{2x}\\
\end{equation}

You can convert this into an operator.

This means this is of the form:
\begin{equation}
\left(\dv[2]{x} - 5\dv{x} + 6\right)y = e^{2x}\\
\end{equation}
Factorise this:
\begin{equation}
\left(\dv{x} - 3\right)\left(\dv{x} - 2\right)y = e^{2x}\\
\end{equation}
If you expand this, you do not operate the operator on the constants -- you multiply it.
So when you expand this for example you do not ie you do not write $\dv{x} \times x = 2\dv{x} \neq 0$.

You now set part of this expression to be equal to $f(x)$ -- say $z$. Now you have reduced the order of the 
differential equation. You solve this as a first order differential equation. Then you get an expression for $z$ 
which you can solve for $z$. Then you solve the remaining system as another linear differential equation.

Let $z = \left(\dv{x} - 2\right)y$.
\begin{equation}
\begin{split}
\left(\dv{x} - 3\right)z &= e^{2x}\\
\dv{z}{x} - 3z &= e^{2x} \\
e^{-3x}\dv{z}{x} - 3ze^{-3x} &= e^{-x} \\
ze^{-3x} &= -e^{-x} + c\\
z &= ce^{3x} - e^{2x} \\
\left(\dv{x} - 2\right)y &= ce^{3x} - e^{2x} \\
\dv{y}{x} - 2y &= ce^{3x} - e^{2x} \\
e^{-2x}\dv{y}{x} - 2ye^{-2x} &= ce^{x} - 1 \\
ye^{-2x} &= ce^{x} - x + d \\
y &= ce^{3x} + de^{2x} - xe^{2x} \\
\end{split}
\end{equation}

This method allows you to solve equations without ever having to make a particular integral -- and allows you to 
solve potentially far more difficult operators. For example one of the differentila operators can 
be a function of $x$ and be near-impossible to be solvable via normal methods.

The definition of $\nabla$ is:
\begin{equation}
\underline{\nabla} =: \underline{i}\pdv{x} + \underline{j}\pdv{y} + \underline{k}\pdv{y} \\
\end{equation}
If this acts on a scalar function $f$, then it is the gradient -- this is why it is called grad.
This represents the normal to any given surface. $\nabla f$ is the normal to any part of that surface.

%Because $\nabla$ is a vector differential operator, it can also operate on other vectors too -- say $\underline{v} = (v_1, v_2, v_3)$.
%This is denoted $\underline{\nabla} \dot \underline{v}$ this is called solenoidial if it is equal to zero.

If $\underline{\nabla}\times \underline{v}$ is equal to zero then it is conserative.
\begin{equation}
\underline{\nabla}\cdot \underline{v} = \pdv{v_1}{x} + \pdv{v_2}{y} + \pdv{v_3}{z} \\
\end{equation}
This is a scalar.

the cross product is equal to the cross product of the nabla operator as a vector and the vector v. 
This gives a vector quantity. 
\begin{equation}
\underline{\nabla}\times\underline{v} = \left(\pdv{v_3}{y} - \pdv{v_2}{z}\right)\underline{i} + \left(\pdv{v_1}{z} - \pdv{v_3}{x}\right)\underline{j} + \left(\pdv{v_1}{y} - \pdv{v_2}{x}\right)\underline{k} \\
\end{equation}

\begin{equation}
\nabla^2 f = \nabla(\nabla f) = \pdv[2]{f}{x} + \pdv[2]{f}{y} + \pdv[2]{f}{z} \\
\end{equation}
This is known as the laplacian and is used in laplaces equation.

\end{document}