\input{../../template/prefs/jp738.tex}
\input{../../template/prefs/me.tex}

\documentclass[10pt,\jkfside,a4paper]{article}

\input{./infofile.tex}
\input{../../template/includes.tex}
\input{../../template/template.tex}

\usepackage{oz}

\begin{document}

\section*{11 On surjections and injections}

\subsection*{11.1 Basic exercises}

\begin{enumerate}

\item Give two examples of functions that are surjective and two examples of functions that are not.



\item give two examples of functions that are injective and two examples of functions that are not.



\end{enumerate}

\subsection*{Core exercises}

\begin{enumerate}

\item Explain and justify the phrase \textit{injections can be undone}



\item Show that $f: A \rightarrow B$ is a surjection if and only if for all sets $C$ and functions $g, : B \rightarrow C$,
$g \circ f = h \circ f$ implies $g = h$.



\end{enumerate}

\section*{12. On images}

\subsection*{Basic exercises}

\begin{enumerate}

\item Let $R_2 = \{(m, n) | m = n^2\}: \mathbb{N} \pfun \mathbb{Z}$ be the integer square-root relation. What is the 
direct image of $\mathbb{N}$ under $R_2$? And what is the inverse image of $\mathbb{N}$.



\item For a relation $R: A \pfun B$, show that:

\begin{enumerate}

\item $\overrightarrow{\text{R}}(X) = \bigcup_{x \in X} \overrightarrow{\text{R}}(\{x\})$ for all $X \subseteq A$.



\item $\overrightarrow{\text{R}}(Y) = \{a \in A | \overrightarrow{\text{R}}(\{a\}) \subseteq Y \}$ for all $Y \subseteq B$.



\end{enumerate}

\end{enumerate}

\subsection*{12.2 Core exercises}

\begin{enumerate}

\item For $X \subseteq A$, prove that the firect image $\overrightarrow{f}(X) \subseteq B$ under an 
injective function $f : A \rightarrowtail B$ is in bijection with $X$, that is, $X \cong \overrightarrow{f}(X)$.



\item Prove that for a surjective function $f: A \twoheadrightarrow B$, the direct image function $\overrightarrow{f}: 
\mathcal{P}(A) \rightarrow \mathcal{P}(B)$ is surjective.



\item Show that any function $f: A \rightarrow B$ can be decomposed into an injection and a surjection: 
that is, there exists a set $X$, a surjection $s: A \twoheadrightarrow X$ and an injection $i: X \rightarrowtail B$ 
such that $f = i \circ s$.



\item For a relation $R: A \pfun B$, prove that

\begin{enumerate}

\item $\overrightarrow{R}(\bigcup \mathcal{F}) = \bigcup \{ \overrightarrow{R}(X) | X \in \mathcal{F}\}$ for all $\mathcal{F} \subseteq \mathcal{P}(X)$



\item $\overleftarrow{R}(\bigcap \mathcal{G}) = \bigcap\{ \overleftarrow{R}(Y) | Y \in \mathcal{G}\}$ for all $\mathcal{G} \subseteq \mathcal{P}(B)$



\end{enumerate}

\item Show that, by the inverse image, every map $A \rightarrow B$ induces a Boolean algebra map $\mathcal{P}(B) \rightarrow \mathcal{P}(A)$. 
That is, for every function $f: A \rightarrow B$, its inverse image preserves set operations:

\begin{itemize}

\item $\overleftarrow{f}(\emptyset) = \emptyset$

\item $\overleftarrow{f}(B) = A$

\item $\overleftarrow{f}(X \cup Y) = \overleftarrow{f}(X) \cup \overleftarrow{f}(Y)$

\item $\overleftarrow{f}(X \cap Y) = \overleftarrow{f}(X) \cap \overleftarrow{f}(Y)$

\item $\overleftarrow{f}(X^\mathsf{c}) = (\overleftarrow{f}(X))^\mathsf{c}$

\end{itemize}

\end{enumerate}

\section*{13 On countability}

\subsection*{13.1 Basic exercises}

\begin{enumerate}

\item Prove that every finite set is countable



\item Demonstrate that $\mathbb{N}, \mathbb{Z}, \mathbb{Q}$ are countable sets.



\end{enumerate}

\subsection*{13.2 Core exercises}

\begin{enumerate}

\item Let $A$ be an infinite subset of $\mathbb{N}$. Show that $A \cong \mathbb{N}$. \\
Hint: Adapt the argument shown in the proof of Proposition 144, showing that the map $\mathbb{N} \rightarrow A$ is both injective and surjective.



\item for an infinite set $A$, prove that the following are equivalent:

\begin{enumerate}

\item There is a bijection $\mathbb{N} \stackrel{\cong}{\rightarrow} A$.

\item There is a surjection $\mathbb{N} \twoheadrightarrow A$.

\item There is an injection $A \rightarrowtail \mathbb{N}$.



\end{enumerate}

\item Prove that:

\begin{enumerate}

\item Every subset of a countable set is countable



\item the product and disjoint union of countable sets is countable.



\end{enumerate}

\item For a set $A$, prove that there is no injection $\mathcal{P}(A) \rightarrowtail A$.



\end{enumerate}

\subsection*{13.3 Optional advanced exercise}

\begin{enumerate}

\item Prove that if $A$ and $B$ are countable sets then so are $A*$, $\mathcal{P}_\text{fin}(A)$ and $\text{PFun}_{\text{fin}}(A, B)$.



\end{enumerate}

\end{document}