\input{./template/prefs/me.tex}

\documentclass[10pt, a4paper]{article}

\input{./template/includes.tex}
\input{./template/template.tex}

\begin{document}

\section*{Machine Learning Supervision 1:}

The ticks do not align with the exam questions -- there is a lot more information in the exam questions and you 
have to get the knowledge from the supervisiosn. The course is quite disjoint.

Anwering the examquestion is a different story.

Read the recommended reading.

Validation and testing data -- get the difference between it.

You are allowed to retest on the validation data -- but on the testing data 
you are only ever allowed to use it once. You must never reuse test data again.
Both are used for evaulation -- but you are allowed to reuse the validation data 
but not test data. Test data must only ever be tested on once.

``What types of words have their sentiments flipped'':
Verbs and ajectives mainly.

Think about ``how is the setup similar to what you have done before'' and how is it different?

When choosing a metric to use you should use a metric which does not allow the system to cheat and get a high score without solving the problem properly. That is how you should analyse metrics.

Baseline approach to a problem:
The simplest approach to a problem which you should compare the model to.
A baseline approach is a "simple and tractible" approach. An approach that you can cod eup in 5 minutes. If by doing something simple you get a very high accuracy then maybe you shoul dnot bother doing anything more complicated. It is a simple and tractible approach you should compare your models with.

When doing (a small number of) binary features we should model the absence of features ie:
\[
P(A\cap B\cap C') = P(A) \times P(B) \times P'(C)
\]

With some features -- like when the features are words -- you should not model the 
absence since the probabilities are so low that the absence of a feature is meaningless.

Burstiness:\\
Words come very infrequently but when they do come they come a lot.

``stop words'' are words which are grammatically necessary.

\end{document}